Circular Regression Trees and Forests with an Application to Probabilistic Wind Direction Forecasting

### 2019-12-20 pkg/circtree/inst/wind_paper/draft/draft.Rnw
Minor spelling corrections and formating
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/partykit@2942 255f4249-2263-4417-b1ea-8498779e27a7

Former-commit-id: 795d4259043e111819d42077e695670f25a38d07

Many approaches to model circular data assume that the circular variable of
interest follows a circular distribution, in particular the von~Mises
distribution which is also known as the ``circular normal distribution''. One
of the first regression models with a circular response variable and linear
covariates was presented by \cite{Gould:1969} where the circular mean
%% , i.e., the location parameter of the von~Mises distribution, 
is predicted by a linear combination of covariates. \cite{Johnson+Wehrly:1978} refined this idea by
plugging in a link function transforming the linear predictor to a restricted
interval of length $2\pi$. This generalized linear model (GLM) type approach
was further extended by \cite{Fisher+Lee:1992} and subsequently by
\cite{Fisher:1993} introducing independent GLMs for either the location or
scale parameter with appropriate link functions while keeping the other
parameter constant. Additionally, they developed a combined heteroscedastic
or distributional version with alternating reestimation of the location and scale parameters
conditional on the respective sets of covariates until convergence. While all of
these models are built on well-elaborated theory, their application in practice
remains very challenging, mainly due to the complexity
encountered in optimizing the corresponding log-likelihood function which is
not globally concave. Therefore, highly-informative starting values are
crucial for such circular GLMs to converge
\citep{Pewsey+Neuhaeuser+Ruxton:2013, Gill+Hangartner:2010}. In order to avoid
this strong dependence on appropriate initial values,
\cite{Mulder+Klugkist:2017} present a Bayesian alternative of a
homoscedastic GLM for circular data. However, apart from potential
difficulties in the optimization procedure of circular GLMs, the interpretation
of the underlying additive effects is often challenging as well because the link function
is highly nonlinear and the representation of smooth transitions on the unit circle
is not straightforward. For example, the same rotation can be
obtained in either positive or negative direction on the circle leading to an
ambiguous interpretation.

@@

Figure~\ref{fig:tree_ibk} illustrates the resulting tree along with
the empirical (gray) and fitted von~Mises (red) wind direction distribution
in each terminal node. Based on the fitted location parameters $\hat \mu$, the
subgroups can be distinguished into the following wind regimes: (1)~Up-valley
winds blowing from the valley mouth towards the upper valley (from east to
west, nodes~4 and 5); (2) Downslope winds blowing across the Alpine crest along
the intersecting valley towards Innsbruck (from south-east to north-west,
node~8); (3) Down-valley winds blowing in the direction of the valley mouth
(from west to east, nodes 10, 12 and 13). Node~7 captures observations with
rather low wind speeds that cannot be clearly distinguished into specific wind
regimes and are consequently associated with a very low estimated concentration
parameter~$\hat{\kappa}$, i.e., a high estimated variance. In terms of
covariates, the lagged wind direction (``persistence'') is mostly responsible
for distinguishing the broad range of wind regimes listed above while the
pressure gradients and wind speed separate the data into subgroups with high
vs.\ low precision. A more extensive case study of circular regression trees
and forests applied to probabilistic wind direction forecasting at
Innsbruck Airport and Vienna International Airport is presented in Section~\ref{sec:wind},
along with a benchmark against commonly-used alternative approaches.

@@

For the specific task of wind forecasting, the wind direction is often only
relevant if the wind speed is sufficiently high. Hence, it is of interest
to account for both quantities simultaneously, e.g., by considering a 
bivariate normal distribution for the wind vectors (from which wind speed
and wind direction can be obtained). The parameters of this bivariate normal
distribution could then be linked to the available covariate using 
an additive regression framework \citep[as proposed by][]{Lang+Mayr+Stauffer:2019}
or using a tree-based approach, similar to the one proposed in this paper.
Moreover, a rather different approach for a combined response of wind speed
and wind direction would be a two-step or hurdle model: In the first step
this could build on the truncated normal model of
\cite{Thorarinsdottir+Gneiting:2010} to capture wind speed. And in a second
step a circular wind direction model is leveraged provided that a certain
hurdle for the wind speed is crossed.

@@

\section*{Computational details}
The corresponding implementation of the proposed methodology for circular
regression trees and forests is provided in the \textsf{R} package \textbf{circtree} 
(version~\Sexpr{packageVersion("circtree")}).
% The proposed circular regression tree and forest models are implemented in the \textsf{R}
% package \textbf{circtree} (version~\Sexpr{packageVersion("circtree")}). 
The package is based on the \textbf{disttree} package
(version~\Sexpr{packageVersion("disttree")}) which applies the main tree
building functions from the \textbf{partykit} package
(version~\Sexpr{packageVersion("partykit")}). All three packages are available
on \textsf{R}-Forge at \url{https://R-Forge.R-project.org/projects/partykit/}.

@@

%\subsection*{CTree}
In the following, the testing and splitting strategy is described for the root node of the tree
which contains the entire learning sample.
%consisting of circular observations denoted by $\{y_i\}_{i = 1,\ldots,n}$, $n \in \mathbb{N}$. 
For a complete tree model, the same procedure is applied iteratively to all resulting child nodes 
with the corresponding subsamples.

First, employing the von~Mises distribution, a distributional model $\text{vM}(y; \mu,\kappa)$ 
is fitted to the learning sample of circular observations $\{y_i\}_{i = 1,\ldots,n}$ as explained 
in Section~\ref{sec:prob_circ}.
In a next step, a goodness-of-fit measurement %of $\text{vM}(Y; \mu,\kappa)$ 
is obtained for each parameter and each observation by evaluating the score 
function $s(\mu, \kappa, y)$ at the estimated location and concentration 
parameter $\hat{\mu}$ and $\hat{\kappa}$.
To detect dependencies between the resulting scoring matrix
\begin{equation}
\begin{pmatrix} 
s(\hat{\mu}, \hat{\kappa}, y_1)_1 \quad, & s(\hat{\mu}, \hat{\kappa}, y_1)_2\\
\vdots & \vdots\\
s(\hat{\mu}, \hat{\kappa}, y_n)_1 \quad,& s(\hat{\mu}, \hat{\kappa}, y_n)_2
\end{pmatrix}
\end{equation}
and each possible split variable $z_l \in \{z_1, \ldots, z_m\}$ a permutation test is applied. 
In particular, the null hypotheses
of independence of each split variable and the scores
%\begin{align}
%H_0^l:  s(\hat{\mu}, \hat{\kappa}, Y) \qquad \bot \qquad Z_l
%\end{align}
is assessed by employing the multivariate linear statistic
\begin{equation}
t_l = vec\left(\sum_{i=1}^n v_l(z_{li}) \cdot s(\hat{\mu}, \hat{\kappa}, y_i)\right)
\end{equation}
with $s(\hat{\mu}, \hat{\kappa}, y_i) \in \mathbb{R}^{1\times 2}$. 
For a numeric split variable $z_l$ the transformation function $v_l$
is simply the identity function $v_l(z_{li}) = z_{li}$ such that $t_l \in \mathbb{R}^2$ 
as the ``$vec$'' operator converts the matrix of dimension $1 \times 2$ into a $2$ column vector.
If $z_l$ is a categorical variable with $h$ categories then 
$v_l(z_{li}) = (\I(z_{li} = 1), \ldots, \I(z_{li} = h))$, hence, $v_l$ returns a unit vector
of dimension $h$ where the entry $1$ indicates the category of $z_{li}$. 
In this case the ``$vec$'' operator converts the $h \times 2$ matrix into a $h \cdot 2$ column vector 
by column-wise combination such that $t_l \in \mathbb{R}^{h \cdot 2}$. 
If there are any observations with missing values these are not included in the calculation of $t_l$.

To map the %observed 
multivariate linear statistic $t_l$ onto the real line a univariate test statistic~$c$ is employed, for example in a quadratic form
\begin{equation}
c_{\text{quad}}(t_l,\mu_l,\Sigma_l) = (t_l-\mu_l)\Sigma_l^+(t_l-\mu_l)^{\top}
\end{equation}
where $\mu_l$ and $\Sigma_l$ are the conditional expectation and the covariance of $t_l$,
as derived by \cite{Strasser+Weber:1999} and used for standardization, and $\Sigma_l^+$ 
is the Moore-Penrose inverse of $\Sigma_l$.
As an alternative, also a maximum form ($c_{\text{max}}$) can be considered such that
the maximum of the absolute values of the standardized linear statistic is returned.

>>>>

Many approaches to model circular data assume that the circular variable of
interest follows a circular distribution, in particular the von~Mises
distribution which is also known as the ``circular normal distribution''. One
of the first regression models with a circular response variable and linear
covariates was presented by \cite{Gould:1969} where the circular mean is
predicted by a linear combination of covariates. \cite{Johnson+Wehrly:1978}
refined this idea by plugging in a link function transforming the linear
predictor to a restricted interval of length $2\pi$. This generalized linear
model (GLM) type approach was further extended by \cite{Fisher+Lee:1992} and
subsequently by \cite{Fisher:1993} introducing independent GLMs for either the
location or scale parameter with appropriate link functions while keeping the
other parameter constant. Additionally, they developed a combined
heteroscedastic or distributional version with alternating reestimation of the
location and scale parameters conditional on the respective sets of covariates
until convergence. While all of these models are built on well-elaborated
theory, their application in practice remains very challenging, mainly due to
the complexity encountered in optimizing the corresponding log-likelihood
function which is not globally concave. Therefore, highly-informative starting
values are crucial for such circular GLMs to converge
\citep{Pewsey+Neuhaeuser+Ruxton:2013, Gill+Hangartner:2010}. In order to avoid
this strong dependence on appropriate initial values,
\cite{Mulder+Klugkist:2017} present a Bayesian alternative of a homoscedastic
GLM for circular data. However, apart from potential difficulties in the
optimization procedure of circular GLMs, the interpretation of the underlying
additive effects is often challenging as well because the link function is
highly nonlinear and the representation of smooth transitions on the unit
circle is not straightforward. For example, the same rotation can be obtained
in either positive or negative direction on the circle leading to an ambiguous
interpretation.

@@

Figure~\ref{fig:tree_ibk} illustrates the resulting tree along with the
empirical (gray) and fitted von~Mises (red) wind direction distribution in each
terminal node. Based on the fitted location parameters $\hat \mu$, the
subgroups can be distinguished into the following wind regimes: (1)~Up-valley
winds blowing from the valley mouth towards the upper valley (from east to
west, nodes~4 and 5); (2)~Downslope winds blowing across the Alpine crest along
the intersecting valley towards Innsbruck (from south-east to north-west,
node~8); (3)~Down-valley winds blowing in the direction of the valley mouth
(from west to east, nodes 10, 12 and~13). Node~7 captures observations with
rather low wind speeds that cannot be clearly distinguished into specific wind
regimes and are consequently associated with a very low estimated concentration
parameter~$\hat{\kappa}$, i.e., a high estimated variance. In terms of
covariates, the lagged wind direction (``persistence'') is mostly responsible
for distinguishing the broad range of wind regimes listed above while the
pressure gradients and wind speed separate the data into subgroups with high
vs.\ low precision. A more extensive case study of circular regression trees
and forests applied to probabilistic wind direction forecasting at Innsbruck
Airport and Vienna International Airport is presented in
Section~\ref{sec:wind}, along with a benchmark against commonly-used
alternative approaches.

@@

For the specific task of wind forecasting, the wind direction is often only
relevant if the wind speed is sufficiently high. Hence, it is of interest to
account for both quantities simultaneously, e.g., by considering a bivariate
normal distribution for wind vectors (from which wind speed and wind direction
can be obtained). The parameters of this bivariate normal distribution could
then be linked to available covariates using an additive regression framework
\citep[as proposed by][]{Lang+Mayr+Stauffer:2019} or using a tree-based
approach, similar to the one proposed in this paper. Moreover, a rather
different approach for a combined response of wind speed and wind direction
would be a two-step or hurdle model: In the first step this could build on the
truncated normal model of \cite{Thorarinsdottir+Gneiting:2010} to capture wind
speed; in the second step a circular wind direction model is leveraged given
that a certain hurdle for the wind speed is crossed.

@@

\section*{Computational details}
The corresponding implementation of the proposed methodology for circular
regression trees and forests is provided in the \textsf{R} package
\textbf{circtree} (version~\Sexpr{packageVersion("circtree")}). The package is
based on the \textbf{disttree} package
(version~\Sexpr{packageVersion("disttree")}) which applies the main tree
building functions from the \textbf{partykit} package
(version~\Sexpr{packageVersion("partykit")}). All three packages are available
on \textsf{R}-Forge at \url{https://R-Forge.R-project.org/projects/partykit/}.

@@

In the following, the testing and splitting strategy is described for the root
node of the tree which contains the entire learning sample. For a complete
tree model, the same procedure is applied iteratively to all resulting child
nodes with the corresponding subsamples.

First, employing the von~Mises distribution, a distributional model
$\text{vM}(y; \mu,\kappa)$ is fitted to the learning sample of circular
observations $\{y_i\}_{i = 1,\ldots,n}$ as explained in
Section~\ref{sec:prob_circ}. In a next step, a goodness-of-fit measurement is
obtained for each parameter and each observation by evaluating the score
function $s(\mu, \kappa, y)$ at the estimated location and concentration
parameter $\hat{\mu}$ and $\hat{\kappa}$. To detect dependencies between the
resulting scoring matrix
\begin{equation}
\begin{pmatrix} 
s(\hat{\mu}, \hat{\kappa}, y_1)_1 \quad, & s(\hat{\mu}, \hat{\kappa}, y_1)_2\\
\vdots & \vdots\\
s(\hat{\mu}, \hat{\kappa}, y_n)_1 \quad,& s(\hat{\mu}, \hat{\kappa}, y_n)_2
\end{pmatrix}
\end{equation}
and each possible split variable $z_l \in \{z_1, \ldots, z_m\}$ a permutation
test is applied. In particular, the null hypotheses of independence of each
split variable and the scores is assessed by employing the multivariate linear
statistic
\begin{equation}
t_l = vec\left(\sum_{i=1}^n v_l(z_{li}) \cdot s(\hat{\mu}, \hat{\kappa}, y_i)\right)
\end{equation}
with $s(\hat{\mu}, \hat{\kappa}, y_i) \in \mathbb{R}^{1\times 2}$. For a
numeric split variable $z_l$ the transformation function $v_l$ is simply the
identity function $v_l(z_{li}) = z_{li}$ such that $t_l \in \mathbb{R}^2$ as
the ``$vec$'' operator converts the matrix of dimension $1 \times 2$ into a $2$
column vector. If $z_l$ is a categorical variable with $h$ categories then
$v_l(z_{li}) = (\I(z_{li} = 1), \ldots, \I(z_{li} = h))$, hence, $v_l$ returns
a unit vector of dimension $h$ where the entry $1$ indicates the category of
$z_{li}$. In this case the ``$vec$'' operator converts the $h \times 2$ matrix
into a $h \cdot 2$ column vector by column-wise combination such that $t_l \in
\mathbb{R}^{h \cdot 2}$. If there are any observations with missing values
these are not included in the calculation of $t_l$.

To map the multivariate linear statistic $t_l$ onto the real line a univariate
test statistic~$c$ is employed, for example in a quadratic form
\begin{equation}
c_{\text{quad}}(t_l,\mu_l,\Sigma_l) = (t_l-\mu_l)\Sigma_l^+(t_l-\mu_l)^{\top}
\end{equation}
where $\mu_l$ and $\Sigma_l$ are the conditional expectation and the covariance
of $t_l$, as derived by \cite{Strasser+Weber:1999} and used for
standardization, and $\Sigma_l^+$ is the Moore-Penrose inverse of $\Sigma_l$.
As an alternative, also a maximum form ($c_{\text{max}}$) can be considered
such that the maximum of the absolute values of the standardized linear
statistic is returned.
